Master thesis

### 2017-06-09 08-Appendix2.Rmd
Appendix2 LL

We gathered data from TRY database to evaluate and improve previous models. We requested functional traits available in Bridge dataset and leaf lifespan for tropical species present in GLOPNET dataset and used in TROLL (see \@ref(tab:A2traits)). We compiled GLOPNET, TRY, and DRYAD databases. We kept dataset given by GLOPNET as origin dataset for observations. It correspond to leaf lifespan and also to LMA and Nmass in most of cases. We measured variable importance in functionnal traits to explain leaf lifespan with out-of bag method applied on a random forest.

@@

```{r functions, echo=FALSE}
fitgraph <- function(fit){
  pars <- fit@model_pars
  beta <- pars[grep('beta', pars)]
  beta <- beta[-grep('s', beta)]
  sigma <- pars[grep('sigma', pars)]
  posterior <- as.matrix(fit)
  r <- mcmc_areas(posterior,  pars = c(beta, sigma, 'lp__'), prob = 0.8)
  c <- mcmc_trace(posterior,  pars = c(beta, sigma, 'lp__'), n_warmup = 2000,
                   facet_args = list(nrow = 2, labeller = label_parsed))
  p <- ggpairs(data.frame(posterior[,beta]))
  return(list(r = r, c = c, p = p))
}
datasub <- function(data, dataset, wsg = TRUE){
  LMAmax <- max(data$LMA)
  Nmax <- max(data$Nm)
  if(wsg)
    wsgmax <- max(data$wsg)
  if(dataset != 'full')
    data <- data[-which(data$Dataset %in% dataset),]
  mdata <- list(S = length(levels(droplevels(data$Dataset))), 
              N = length(as.numeric(droplevels(data$Dataset))),
              J = as.numeric(droplevels(data$Dataset)),
              LL = data$LL,
              LMA = (data$LMA/LMAmax),
              Nmass = (data$Nm/Nmax))
  if(wsg)
    mdata$wsg <- data$wsg/wsgmax
  return(mdata)
}
```

We tested different models and we kept the model with the best tradeoff between model complexity, convergence, likelihood, and prediction quality (see [leaf lifespan model](LL2.html) document). We found following model, named **M1**:
$$M1~~ {LL_s}_j \sim \mathcal{logN}(\beta_0*e^{{\beta_1}_s*{LMA_s}_j - {\beta_2}_s*{Nmass_s}_j},\,\sigma)\,$$
$$s=1,...,S_{=4}~, ~~j=1,...,n_s$$
$${\beta_i}_s \sim \mathcal{N}({\beta_i},\,\sigma_i)\,^I, ~~(\beta_i, \sigma, \sigma_i) \sim \mathcal{\Gamma}(0.001,\,0.001)\,^{2I+1}$$

Leaf lifespan ($LL$) is following a lognormal law with location ${\mu_s}_j$ and scale $\sigma$. $s$ represents the dataset used to do the fit between 1 and $S=4$, it encompass environmental and protocol variations. $j$ represents the observation for a given site $s$. Location ${\mu_s}_j$ is the exponential from the difference between LMA and Nmass with parameters ${\beta_0}_s$, ${\beta_1}_s$, and ${\beta_2}_s$. Each ${\beta_i}_s$ is following a normal law located on $\beta_i$ with scale $\sigma_i$. Finally all $beta_i$, $\sigma$, and $\sigma_i$ are assumed without presemption following gamma laws of parameters $10^-2$, $10^-2$.

We decided to add to model **M1**, wood specific gravity as a new predictor through a parameter $\beta_3$ in a new model named **M10**. We removed dataset random effect on $\beta_3$ due to the fact that all wood specific gravities were provided by DRYAD database. Due to a correlation between $\beta_0$ and $\beta_3$ in **M10** (see figure 5), we tried also a model **M11** without intercept $\beta_0$. We then sampled all models with this composite dataset, and obtained following results :

@@

```{r OOB, echo=FALSE}
knitr::kable(randomForest::randomForest(LL ~ LMA + Nm + LMA + wsg,
                   data = data, na.action = na.omit,
                   ntree = 1000, replace = F , importance = T,
                   do.trace = F, keep.forest = T, keep.inbag = T)$importance,
             caption = 'Variable importance calculated with out-of the bag method applied on a random forest. Leaf lifespan (LL) is taken in GLOPNET database from @wright_worldwide_2004. Leaf mass per area (LMA), and leaf nitrogen content (Nmass) are taken both in TRY (https://www.try-db.org) and GLOPNET databases. Wood specific gravity (wsg) is taken both in TRY and DRYAD databases.', format = 'pandoc')
```

Table 3: Model fit results for composite dataset. *Data are taken from GLOPNET, TRY, and DRYAD datasets (see previous description). ML stand for maximum likelihood and RMSEP for root mean square error of prediction*

```{r model, echo=FALSE}
# mdata <- sapply(c('full', levels(droplevels(data$Dataset))),
#                 function(x) datasub(data, x), simplify = F)
# fits <- lapply(as.list(paste0('M', c(1, 10, 11))), function(model){
#   fit <- lapply(mdata, function(x) stan(file = file.path(mpath, paste0(model, '.stan')), data = x))
#   eval <- NA
#   fit <- list(full = fit$full, eval = eval)
#   path <- file.path(mpath, paste0(model, '.LL.Rdata'))
#   save(fit, file = path)
# })
# fits <- list()
# i <- 1
# models <- list.files(path = mpath, pattern = '.LL.Rdata')
# for(file in models){
#   load(file = file.path(mpath, file))
#   fits[[i]] <- fit ; rm(fit) ; i <- i + 1
# }
# names(fits) <- paste0('fit', c(10, 11, 1))
# g10 <- fitgraph(fits$fit10$full)
# g11 <- fitgraph(fits$fit11$full) ; rm(file, i, models, mpath)
# tab <- data.frame(
#   M = c('M1', 'M10', 'M11'),
#   Model = c('${LL_s}_j \\sim \\mathcal{logN}(\\beta_0 + {\\beta_1}_s*{LMA_s}_j - {\\beta_2}_s*{Nmass_s}_j,\\,\\sigma)\\,$',
#             '${LL_s}_j \\sim \\mathcal{logN}(\\beta_0 + {\\beta_1}_s*{LMA_s}_j - {\\beta_2}_s*{Nmass_s}_j + {\\beta_3}*{wsg}_j,\\,\\sigma)\\,$',
#             '${LL_s}_j \\sim \\mathcal{logN}({\\beta_1}_s*{LMA_s}_j - {\\beta_2}_s*{Nmass_s}_j + {\\beta_3}*{wsg}_j,\\,\\sigma)\\,$'),
#   ML = c( max(as.matrix(fits$fit1$full)[,'lp__']),
#           max(as.matrix(fits$fit10$full)[,'lp__']),
#           max(as.matrix(fits$fit11$full)[,'lp__']))
# )
# kable(tab)
```

All model have correctly converged.

```{r A2r, echo=FALSE, fig.cap='Parameters posterior ditribution. Light blue area represents 80% confidence interval, and vertical blue line the mean.'}
# g11$r
```

```{r A2c, echo=FALSE, fig.cap='Parameters markov chains. Light grey area represents warmup iterations.'}
# g11$c
```

Model **M10** have shown a correlation between parameters $\beta_0$ and $\beta_3$, responsible for the creation of model **M11** (see previous paragraphs). Model **M11** only show a slight link between $\beta_1$ and $\beta_3$ which can be considered as non significant.

```{r A2p11, echo=FALSE, fig.cap='Parameters pairs plot for model M11. Parameters density distribution, pairs plot and Pearson s coefficient of correlation.'}
# g11$p + ggtitle('M11') 
```

We can denote that retained vector of parameters with maximum likelihood for model **M11** is close to 95% confidence limit.

```{r A2graph11, echo=FALSE, fig.cap='Leaflifespan predictions for model M11 with leaf mass per area (LMA), leaf nitrogen content (Nmass), wood specific gravity (wsg) and predicted versus observed values. Leaf lifespan (LL) is predicted with model M10 fit. Leaf mass per area (LMA) and leaf nitrogen content (Nmass), and wood specific gravity (wsg) are taken in a composite dataset of GLOPNET, TRY and DRYAD  datasets.Warning LMA (resp. Nmass and wsg) is not constant and depend on the closest point value for right (resp. center and left) graph.'}
# beta <- as.matrix(fits$fit11$full)[which.max(as.matrix(fits$fit11$full)[,'lp__']),c('beta_1', 'beta_2', 'beta_3')]
# fit <- fits$fit10$full
# pred <- apply(as.matrix(fit), 1, function(x)
#   sapply(seq_len(dim(data)[1]), function(i)
#       rlnorm(1, x['beta_1']*(data$LMA[i]/LMAmax) - x['beta_2']*(data$Nm[i]/Nmax) + x['beta_3']*(data$wsg[i]/wsgmax), x['sigma'])
#     ))
# pred <- data.frame(t(apply(pred, 1, function(x) quantile(x, probs = c(0.15, 0.85)))))
# names(pred) <- c('5%', '95%')
# pred$mean <- exp((beta[1]*(data$LMA/LMAmax) - beta[2]*(data$Nm/Nmax) + beta[3]*(data$wsg/wsgmax)))
# pred$LMA <- data$LMA
# pred$Nm <- data$Nm
# pred$wsg <- data$wsg
# pred$obs <- data$LL
# lma <- ggplot(data, aes(x = LMA, y = LL)) +
#   geom_ribbon(aes(x = pred$LMA, ymin = pred$`5%`, ymax = pred$`95%`), color = 'grey', alpha = 0.2) +
#   geom_line(aes(x = pred$LMA, y = pred$`5%`, color = '5%'), linetype = 2) +
#   geom_line(aes(x = pred$LMA, y = pred$`95%`, color = '95%'), linetype = 2) +
#   geom_line(aes(x = pred$LMA, y = pred$mean, color = 'mean'), linetype = 2) +
#   geom_point() +
#   theme_bw() +
#   xlab('Leaf mass per area (LMA in g/m2)') +
#   ylab('Leaf lifespan (LL in months)') +
#   scale_color_manual(name = 'Model',
#                      values = c('5%' = 'black',
#                                 '95%' = 'black',
#                                 'mean' = 'red'))
# nm <- ggplot(data, aes(x = Nm, y = LL)) +
#   geom_ribbon(aes(x = pred$Nm, ymin = pred$`5%`, ymax = pred$`95%`), color = 'grey', alpha = 0.2) +
#   geom_line(aes(x = pred$Nm, y = pred$`5%`, color = '5%'), linetype = 2) +
#   geom_line(aes(x = pred$Nm, y = pred$`95%`, color = '95%'), linetype = 2) +
#   geom_line(aes(x = pred$Nm, y = pred$mean, color = 'mean'), linetype = 2) +
#   geom_point() +
#   theme_bw() +
#   xlab('Leaf nitrogen content (Nmass in %)') +
#   ylab('Leaf lifespan (LL in months)') +
#   scale_color_manual(name = 'Model',
#                      values = c('5%' = 'black',
#                                 '95%' = 'black',
#                                 'mean' = 'red'))
# wsg<- ggplot(data, aes(x = wsg, y = LL)) +
#   geom_ribbon(aes(x = pred$wsg, ymin = pred$`5%`, ymax = pred$`95%`), color = 'grey', alpha = 0.2) +
#   geom_line(aes(x = pred$wsg, y = pred$`5%`, color = '5%'), linetype = 2) +
#   geom_line(aes(x = pred$wsg, y = pred$`95%`, color = '95%'), linetype = 2) +
#   geom_line(aes(x = pred$wsg, y = pred$mean, color = 'mean'), linetype = 2) +
#   geom_point() +
#   theme_bw() +
#   xlab('Leaf nitrogen content (Nmass in %)') +
#   ylab('Leaf lifespan (LL in months)') +
#   scale_color_manual(name = 'Model',
#                      values = c('5%' = 'black',
#                                 '95%' = 'black',
#                                 'mean' = 'red'))
# predvsobs<- ggplot(pred, aes(x = obs, y = mean)) +
#   geom_ribbon(aes(ymin =`5%`, ymax = `95%`), color = 'grey', alpha = 0.2) +
#   geom_line(aes(y = `5%`, color = '5%'), linetype = 2) +
#   geom_line(aes(y = `95%`, color = '95%'), linetype = 2) +
#   geom_line(aes(y = mean, color = 'mean'), linetype = 2) +
#   geom_point() +
#   theme_bw() +
#   xlab('Observed leaf lifespan (LL in months)') + 
#   ylab('Predicted leaf lifespan (LL in months)') +
#   geom_abline(slope = 1) +
#   scale_color_manual(name = 'Model',
#                      values = c('5%' = 'black',
#                                 '95%' = 'black',
#                                 'mean' = 'red'))
# plot_grid(lma, nm, wsg, predvsobs, labels = c('LMA', 'Nmass', 'wsg', 'P v. O'))

@@

We finally tested the new allometry obtained with the new model **M** on TROLL data (figure 7) and with TROLL simulations. Minimum leaf lifespan is now above 3 months for the new model **M** with TROLL species dataset. Moreover lowest leaf lifespan species are not the one with the smallest LMA with new model. Lastly, new model **M** variation of leaf lifespan is almost spanning all previous allometries prediciton on LMA axis.

```{r TROLL, echo=FALSE, fig.cap='Leaflifespan predictions for model different allometries with leaf mass per area (LMA), and leaf nitrogen content (Nmass), and wood specific gravity (wsg). Leaf lifespan (LL) is predicted with model M1, M10, and M11 fits or Reich s and Wright s allometries [@Reich1991a]. Leaf mass per area (LMA) leaf nitrogen content (Nmass) , and wood specific gravity (wsg) are taken in BRIDGE dataset used by TROLL [@Li].'}
# species <- read.table('~/Documents/ECOFOG/TROLL/inst/extdata/species.txt', header=T, dec=".", sep="", row.names = 1)[c('LMA', 'Nmass', 'wsg')]
# species$`Reich 91` <- 1.5 + 10^(7.18+3.03*log10(species$LMA*0.0001))
# species$`Reich 97` <- 10^(2.040816*(2.579713-log10(10000.0/species$LMA)))
# species$`Wright 04` <- 0.5+10^(-2.509+1.71*log10(species$LMA))
# beta <- as.matrix(fits$fit1$full)[which.max(as.matrix(fits$fit1$full)[,'lp__']),c('beta_0', 'beta_1', 'beta_2')]
# species$Model1 <- with(species, exp((beta[1] + beta[2]*(LMA/LMAmax) - beta[3]*(Nmass*100/Nmax))))
# beta <- as.matrix(fits$fit10$full)[which.max(as.matrix(fits$fit10$full)[,'lp__']),c('beta_0', 'beta_1', 'beta_2', 'beta_3')]
# species$Model10 <- with(species, exp((beta[1] + beta[2]*(LMA/LMAmax) - beta[3]*(Nmass*100/Nmax) + beta[4]*(wsg/wsgmax))))
# beta <- as.matrix(fits$fit11$full)[which.max(as.matrix(fits$fit11$full)[,'lp__']),c('beta_1', 'beta_2', 'beta_3')]
# species$Model11 <- with(species, exp((beta[1]*(LMA/LMAmax) - beta[2]*(Nmass*100/Nmax) + beta[3]*(wsg/wsgmax))))
# data <- melt(species, id = c('LMA', 'Nmass', 'wsg'))
# names(data)[4] <- 'Allometry'
# g <- ggplot(data, aes(x = LMA, y = value, color = Allometry)) +
#   geom_point() + theme_bw() +   xlab('LMA in g/m2') +
#   ylab('Leaf lifespan (LL in months)')
# h <- ggplot(data, aes(x = Nmass, y = value, color = Allometry)) +
#   geom_point() + theme_bw() +   xlab('Nmass in %') +
#   ylab('Leaf lifespan (LL in months)')
# f <- ggplot(data, aes(x = wsg, y = value, color = Allometry)) +
#   geom_point() + theme_bw() +   xlab('wsg in mg/mm3') +
#   ylab('Leaf lifespan (LL in months)')
# plot_grid(g, h, f, labels = c('LMA', 'Nmass', 'wsg'))
```

Simulations are validating that this new allometry resolve the issue of unrealistically low aboveground biomass for low LMA species due to an early edath of individuals inside simulations. For instance with this allometry Symphonia sp 1 (a low LMA species) is now reaching a realistic aboveground biomass above X tonnes of C/ha and realistic diameter and age distribution inside the final population.

>>>>

We compiled functional traits from GLOPNET [@wright_worldwide_2004], TRY [@Kattge2011], and DRYAD [@chave_towards_2009] databases (see \@ref(tab:A2traits)). We kept dataset given by GLOPNET as origin dataset for observations. Dataset defined as origin corresponded to leaf lifespan ($LL$) and most of the time to leaf mass per area ($LMA$) and leaf nitrogen content per leaf dry mass ($Nmass$). We measured variable importance in functionnal traits to explain leaf lifespan with out-of bag method applied on a random forest. Then, we used a béayesian apporach to test different models with growing level of complexity. We retained the model with the best tradeoff between model complexity (number of parameters $K$), convergence, likelihood, and prediction quality (root mean sqaure error of prediciton $RMSEP$). We finally tested the new allometry obtained with the selected model with TROLL simulations.

@@

@@

Out of the bag method applied on a random forest highlighted the importance of leaf nitrogen content per leaf dry mass ($N_m$) to model leaf lifespan (see \@ref(tab:A2OOB)). $N_m$ importance was higher than leaf mass per area (158 against 96 percent of mean square error increase) which was used as a proxy for leaf lifespan in previous models. Finally, wood specific gravity ($wsg$) add also a significant importance in leaf lifespan estimation.

```{r A2OOB, echo=FALSE}
knitr::kable(randomForest::randomForest(LL ~ LMA + Nm + LMA + wsg,
                   data = data, na.action = na.omit,
                   ntree = 1000, replace = F , importance = T,
                   do.trace = F, keep.forest = T, keep.inbag = T)$importance,
             caption = 'Variable importance calculated with out-of the bag method applied on a random forest. First column represents the mean decrease in mean square error (%IncMSE) whereas second column represents the total decrease in node impurities, measured by the Gini Index  (IncNodePurety). Leaf lifespan (LL) is taken in GLOPNET database from @wright_worldwide_2004. Leaf mass per area (LMA), and leaf nitrogen content (Nmass) are taken both in TRY (https://www.try-db.org) and GLOPNET databases. Wood specific gravity (wsg) is taken both in TRY and DRYAD databases.', format = 'pandoc')
```

The selected model had a maximum likelihood of 13.6 and a RMSEP of 12 months:

\begin{equation}
  LL_{d} \sim log\mathcal{N}({\beta_1}_d*LMA - {\beta_2}_d*N*\beta_3*wsg, \sigma)
  (\#eq:LL)
\end{equation}

Leaf lifespan $LL$ follows a lognormal law with location infered from leaf lifespan $LMA$, nitrogen content $N$ and wood specific gravity $wsg$ and a scale $\sigma$. Each ${\beta_i}_d$ is following a normal law located on $\beta_i$ with a scale of $\sigma_i$. All $\beta_i$, $\sigma_i$, and $\sigma$ are assumed without presemption following a gamma law. $d$ represents the dataset random effects and encompass environmental and protocol variations.

```{r A2LLpred, echo=FALSE, fig.cap='Leaflifespan predictions for the selected model with leaf mass per area (LMA), leaf nitrogen content (Nmass), wood specific gravity (wsg) and predicted versus observed values. Leaf lifespan (LL) is predicted with model M10 fit. Leaf mass per area (LMA) and leaf nitrogen content (Nmass), and wood specific gravity (wsg) are taken in a composite dataset of GLOPNET, TRY and DRYAD  datasets.Warning LMA (resp. Nmass and wsg) is not constant and depend on the closest point value for right (resp. center and left) graph.', fig.height=12, fig.width=12}
library(ggplot2)
load("~/Documents/ECOFOG/TROLL/inst/doc/LL_models/M11.LL.Rdata")
fit <- fit$full
beta <- as.matrix(fit)[which.max(as.matrix(fit)[,'lp__']),c('beta_1', 'beta_2', 'beta_3')]
pred <- apply(as.matrix(fit), 1, function(x)
  sapply(seq_len(dim(data)[1]), function(i)
      rlnorm(1, x['beta_1']*(data$LMA[i]/LMAmax) - x['beta_2']*(data$Nm[i]/Nmax) + x['beta_3']*(data$wsg[i]/wsgmax), x['sigma'])
    ))
pred <- data.frame(t(apply(pred, 1, function(x) quantile(x, probs = c(0.15, 0.85)))))
names(pred) <- c('5%', '95%')
pred$mean <- exp((beta[1]*(data$LMA/LMAmax) - beta[2]*(data$Nm/Nmax) + beta[3]*(data$wsg/wsgmax)))
pred$LMA <- data$LMA
pred$Nm <- data$Nm
pred$wsg <- data$wsg
pred$obs <- data$LL
lma <- ggplot(data, aes(x = LMA, y = LL)) +
  geom_ribbon(aes(x = pred$LMA, ymin = pred$`5%`, ymax = pred$`95%`), color = 'grey', alpha = 0.2) +
  geom_line(aes(x = pred$LMA, y = pred$`5%`, color = '5%'), linetype = 2) +
  geom_line(aes(x = pred$LMA, y = pred$`95%`, color = '95%'), linetype = 2) +
  geom_line(aes(x = pred$LMA, y = pred$mean, color = 'mean'), linetype = 2) +
  geom_point() +
  theme_bw() +
  xlab('Leaf mass per area (LMA in g/m2)') +
  ylab('Leaf lifespan (LL in months)') +
  scale_color_manual(name = 'Model',
                     values = c('5%' = 'black',
                                '95%' = 'black',
                                'mean' = 'red'))
nm <- ggplot(data, aes(x = Nm, y = LL)) +
  geom_ribbon(aes(x = pred$Nm, ymin = pred$`5%`, ymax = pred$`95%`), color = 'grey', alpha = 0.2) +
  geom_line(aes(x = pred$Nm, y = pred$`5%`, color = '5%'), linetype = 2) +
  geom_line(aes(x = pred$Nm, y = pred$`95%`, color = '95%'), linetype = 2) +
  geom_line(aes(x = pred$Nm, y = pred$mean, color = 'mean'), linetype = 2) +
  geom_point() +
  theme_bw() +
  xlab('Leaf nitrogen content (Nmass in %)') +
  ylab('Leaf lifespan (LL in months)') +
  scale_color_manual(name = 'Model',
                     values = c('5%' = 'black',
                                '95%' = 'black',
                                'mean' = 'red'))
wsg<- ggplot(data, aes(x = wsg, y = LL)) +
  geom_ribbon(aes(x = pred$wsg, ymin = pred$`5%`, ymax = pred$`95%`), color = 'grey', alpha = 0.2) +
  geom_line(aes(x = pred$wsg, y = pred$`5%`, color = '5%'), linetype = 2) +
  geom_line(aes(x = pred$wsg, y = pred$`95%`, color = '95%'), linetype = 2) +
  geom_line(aes(x = pred$wsg, y = pred$mean, color = 'mean'), linetype = 2) +
  geom_point() +
  theme_bw() +
  xlab('Leaf nitrogen content (Nmass in %)') +
  ylab('Leaf lifespan (LL in months)') +
  scale_color_manual(name = 'Model',
                     values = c('5%' = 'black',
                                '95%' = 'black',
                                'mean' = 'red'))
predvsobs<- ggplot(pred, aes(x = obs, y = mean)) +
  geom_ribbon(aes(ymin =`5%`, ymax = `95%`), color = 'grey', alpha = 0.2) +
  geom_line(aes(y = `5%`, color = '5%'), linetype = 2) +
  geom_line(aes(y = `95%`, color = '95%'), linetype = 2) +
  geom_line(aes(y = mean, color = 'mean'), linetype = 2) +
  geom_point() +
  theme_bw() +
  xlab('Observed leaf lifespan (LL in months)') +
  ylab('Predicted leaf lifespan (LL in months)') +
  geom_abline(slope = 1) +
  scale_color_manual(name = 'Model',
                     values = c('5%' = 'black',
                                '95%' = 'black',
                                'mean' = 'red'))
cowplot::plot_grid(lma, nm, wsg, predvsobs, labels = c('LMA', 'Nmass', 'wsg', 'P v. O'))

@@

Simulations are validating that this new allometry resolve the issue of unrealistically low aboveground biomass for low LMA species due to an early edath of individuals inside simulations. For instance with this allometry Symphonia sp 1 (a low LMA species) is now reaching a realistic aboveground biomass above 400 $tonC.ha^{-1}$ and realistic diameter and age distribution inside the final population.
